A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-14 paper.Rmd
section 2.3 but needs more work

The examination of data plots to infer signal or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in graph [@cleveland_graphical_1984]. Such examination thus feels there is a subjective nature to evaluate, say, diagnostic plots to infer appropriateness of statistical models, although arguably we would expect such evaluation should be done objectively and such human-to-human variation in "reading" data plots should be minimised.

In practice, over-interpretation of a single plot is common.  For instance, in @roy_chowdhury_using_2015, some over-interpreted the separation between gene groups in a two-dimensional projection of a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups. To mitigate this over-interpretation, @buja_statistical_2009 proposed a line-up protocol to assess plots in a manner analogous to the null hypothesis significance testing (NHST) framework. More specifically, the protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, have the identical graphical procedure as the data plot except the data is replaced with a data generation mechanism that is consistent with the null hypothesis, $H_0$. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires apriori specification of $H_0$ (or at least a null data generating mechanism) much like the requirement of knowing the distribution of the test statistic in NHST.

>>>>

The examination of data plots to infer signal or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in a graph [@cleveland_graphical_1984]. To rely on human examination thus feels as though there is a subjective nature to using diagnostic plots to infer appropriateness of statistical models. Ideally we would expect visual evaluation should be done objectively and human-to-human variation in reading data plots small, but it cannot be assured.

In practice, over-interpretation of a single plot is common.  For instance, in @roy_chowdhury_using_2015, described an published example where the authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis. @roy_chowdhury_using_2015 showed that there were no differences in the expression levels between the gene groups. The lineup protocol proposed in @buja_statistical_2009 provides a solution to over-interpretation.  @majumder_validation_2013 showed that its use is analogous to the null hypothesis significance testing (NHST) framework. More specifically, the protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, have the identical graphical procedure as the data plot except the data is replaced with a data generation mechanism that is consistent with the null hypothesis, $H_0$. Then, an observer who have not seen the data plot will be asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires apriori specification of $H_0$ (or at least a null data generating mechanism) much like the requirement of knowing the distribution of the test statistic in NHST.

### 2023-04-14 paper.tex
section 2.3 but needs more work

The examination of data plots to infer signal or patterns (or lack
thereof) is fraught with variation in the human ability to interpret and
decode the information embedded in graph
\citep{cleveland_graphical_1984}. Such examination thus feels there is a
subjective nature to evaluate, say, diagnostic plots to infer
appropriateness of statistical models, although arguably we would expect
such evaluation should be done objectively and such human-to-human
variation in ``reading'' data plots should be minimised.

In practice, over-interpretation of a single plot is common. For
instance, in \citet{roy_chowdhury_using_2015}, some over-interpreted the
separation between gene groups in a two-dimensional projection of a
linear discriminant analysis when in fact there are no differences in
the expression levels between the gene groups. To mitigate this
over-interpretation, \citet{buja_statistical_2009} proposed a line-up
protocol to assess plots in a manner analogous to the null hypothesis
significance testing (NHST) framework. More specifically, the protocol
consists of \(m\) randomly placed plots, where one plot is the data
plot, and the remaining \(m - 1\) plots, referred to as the \emph{null
plots}, have the identical graphical procedure as the data plot except
the data is replaced with a data generation mechanism that is consistent
with the null hypothesis, \(H_0\). Then, an observer who have not seen
the data plot will be asked to point out the most different plot from
the lineup. Under \(H_0\), it is expected that the data plot would have
no distinguishable difference from the null plots, and the probability
that the observer correctly picks the data plot is \(1/m\). If one
rejects \(H_0\) as the observer correctly picks the data plot, then the
Type I error of this test is \(1/m\). This protocol requires apriori
specification of \(H_0\) (or at least a null data generating mechanism)
much like the requirement of knowing the distribution of the test
statistic in NHST.

>>>>

The examination of data plots to infer signal or patterns (or lack
thereof) is fraught with variation in the human ability to interpret and
decode the information embedded in a graph
\citep{cleveland_graphical_1984}. To rely on human examination thus
feels as though there is a subjective nature to using diagnostic plots
to infer appropriateness of statistical models. Ideally we would expect
visual evaluation should be done objectively and human-to-human
variation in reading data plots small, but it cannot be assured.

In practice, over-interpretation of a single plot is common. For
instance, in \citet{roy_chowdhury_using_2015}, described an published
example where the authors over-interpreted separation between gene
groups from a two-dimensional projection of a linear discriminant
analysis. \citet{roy_chowdhury_using_2015} showed that there were no
differences in the expression levels between the gene groups. The lineup
protocol proposed in \citet{buja_statistical_2009} provides a solution
to over-interpretation. \citet{majumder_validation_2013} showed that its
use is analogous to the null hypothesis significance testing (NHST)
framework. More specifically, the protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) plots, referred to as the \emph{null plots}, have the
identical graphical procedure as the data plot except the data is
replaced with a data generation mechanism that is consistent with the
null hypothesis, \(H_0\). Then, an observer who have not seen the data
plot will be asked to point out the most different plot from the lineup.
Under \(H_0\), it is expected that the data plot would have no
distinguishable difference from the null plots, and the probability that
the observer correctly picks the data plot is \(1/m\). If one rejects
\(H_0\) as the observer correctly picks the data plot, then the Type I
error of this test is \(1/m\). This protocol requires apriori
specification of \(H_0\) (or at least a null data generating mechanism)
much like the requirement of knowing the distribution of the test
statistic in NHST.
